On the propriety of restricted Boltzmann machines

### 2016-11-05 writing/appendix/appendix-center.Rmd
reorganize dir and try to soften language about empirical likelihood

1. The function $\bm{g}: \mathbb{R}^{m+n+mn} \rightarrow \mathcal{I}$ is one-to-one; see Theorem \ref{thm:one} below. Hence, these moments uniquely characterize the parameters in the RBM$(m,n,\bm{\theta})$ model. (Additionally, two RBM models have the same parameters if and only if the models have the same probability functions.)
1. The image $\mathcal{I}$ is a symmetric set (i.e., $\bm{g}(\bm{\theta})=-\bm{g}(-\bm{\theta})$) and must be a connected (not necessarily convex) region in $\mathbb{R}^{m+n+mn}$, without any voids in it (by the continuity of $\bm{g}(\bm{\theta})$).
1. The center of the image $\mathcal{I}$ (the zero vector in $\mathbb{R}^{m+n+mn}$) is given only by $\bm{\theta}=\bm{0}$ (the zero vector in the parameter space).  This corresponds to the case where all the random vectors $V_1,\ldots,V_m,H_1,\ldots,H_n$ are iid with a symmetric Bernoulli$(1/2)$ distribution; see Theorem \ref{thm:two}.

@@

{\bf Proof of Theorem \ref{thm:one}.} We first establish the equivalence of claims 2 and 3. If claim 2 holds, then
 claim 3 holds trivially. Suppose claim 3 holds and pick $\ell \in \{1,\ldots,m\}$. Let  $\theta_{v_1},\ldots,\theta_{v_m},\theta_{h_1},\ldots,\theta_{h_n}$, $\theta_{ij}$, $i=1,\ldots,m$, $j=1,\ldots,n$, denote the components of $\bm{\theta}_1$ and let $\tilde{\theta}_{v_1},\ldots,\tilde{\theta}_{v_m},\tilde{\theta}_{h_1},\ldots,\tilde{\theta}_{h_n}$, $\tilde{\theta}_{ij}$, $i=1,\ldots,m$, $j=1,\ldots,n$, denote the analogous components of $\bm{\theta}_2$. Then, by assumption,
\begin{align*}
\exp\left[ 2\theta_{v_\ell} +2 \sum_{j=1}^n  h_j \theta_{\ell j} \right] &= \frac{P_1\left(\{V_\ell = 1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)} {P_1\left(\{V_\ell = -1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}\\
&= \frac{P_2\left(\{V_\ell = 1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}
{P_2\left(\{V_\ell = -1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}\exp\left[ 2\tilde{\theta}_{v_\ell} +2 \sum_{j=1}^n  h_j \tilde{\theta}_{\ell j} \right]
\end{align*}
for any $\{v_i\}_{i=1,i\neq \ell}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\}$. Equating exponents, we have $\theta_{v_\ell} - \tilde{\theta}_{v_\ell} = \sum_{j=1}^n  h_j (  \tilde{\theta}_{\ell j} -\theta_{\ell j})$ for any $\{h_j\}_{j=1}^n \subset \{\pm 1\}$. Taking all $h_j=1$ or all $h_j=-1$ for $1 \leq j \leq n$ yields $\theta_{v_\ell} = \tilde{\theta}_{v_\ell}$. This result further implies $\tilde{\theta}_{\ell 1} -\theta{\ell 1} = -  \sum_{j=2}^n  h_j (  \tilde{\theta}_{\ell j} -\theta_{\ell j})$ for
any given $\{h_j\}_{j=2}^n \subset \{\pm 1\}$, so that $\theta_{\ell 1}=\tilde{\theta}_{\ell 1}$ must hold. Iterating this argument, sequentially for each $j=2,\ldots,m$, shows $\theta_{\ell j}=\tilde{\theta}_{\ell j}$ for any $1 \leq j \leq n$ in addition to $\theta_{v_\ell}=\tilde{\theta}_{v_\ell}$.  As $\ell\in\{1,\ldots,m\}$ was arbitrary, we have $\theta_{v_i}=\tilde{\theta}_{v_i}$ and $\theta{i j}=\tilde{\theta}_{i j}$ for any $1 \leq i \leq m$, $1 \leq j \leq n$.

Now pick $k \in \{1,\ldots,n\}$ so that, by the RBM$(m,n,\bm{\theta})$ probability structure as above, we have analogously that
$$
\theta_{h_k} + \sum_{i=1}^m  v_i \theta_{i k} =  \tilde{\theta}_{h_k} + \sum_{i=1}^m  v_i \tilde{\theta}_{i k}
$$
fsor any $\{v_i\}_{i=1}^m \subset \{\pm 1\}$, so that $\theta_{h_k}  =  \tilde{\theta}_{h_k}$. As $k \in \{1,\ldots,n\}$ was arbitrary, we have now established claim 2 from claim 3.

To establish the equivalence of claims 1 and 3 in Theorem \ref{thm:one} under the RBM$(m,n,\bm{\theta})$ model, we note that claim 3 easily implies claim 1. Hence, it suffices now to establish claim 3 from claim 1. This follows immediately from Lemma \ref{lemma:one} below and the fact that the variables $V_1,\ldots,V_m$ are independent given $H_1,\ldots,H_n$ in the RBM$(m,n,\bm{\theta})$ model (and likewise $H_1,\ldots,H_n$ are independent given $V_1,\ldots,V_m$).

\begin{applemma} 
\label{lemma:one}
Let $m,n\geq 1$.  Let $P_1,P_2$ denote two probability distributions for a random vector $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ supported on $\{a,b\}^{m+n}$, for some real constants $a\neq b \in\mathbb{R}$, such that $P_i( \bm{V}_m=\bm{v}_m, \bm{H}_n=\bm{h}_n)>0$, $i=1,2$, for any $\bm{v}_m \in \{a,b\}^{m}$, $\bm{h}_n \in \{a,b\}^{n}$. Suppose additionally that, under $P_1$ or $P_2$,  the variables $V_1,\ldots,V_m$ are independent given $\bm{H}_n=\bm{h}_n \in \{a,b\}^{n}$ and that the variables $H_1,\ldots,H_n$ are independent given $\bm{V}_m=\bm{v}_m \in \{a,b\}^{m}$.

>>>>

\begin{itemize}
\item The function $\bm{g}: \mathbb{R}^{m+n+mn} \rightarrow \mathcal{I}$ is one-to-one; see Theorem \ref{thm:one} below. Hence, these moments uniquely characterize the parameters in the RBM$(m,n,\bm{\theta})$ model.
\item The image $\mathcal{I}$ is a symmetric set (i.e., $\bm{g}(\bm{\theta})=-\bm{g}(-\bm{\theta})$) and must be a connected (not necessarily convex) region in $\mathbb{R}^{m+n+mn}$, without any voids in it (by the continuity of $\bm{g}(\bm{\theta})$).
\item The center of the image $\mathcal{I}$ (the zero vector in $\mathbb{R}^{m+n+mn}$) is given only by $\bm{\theta}=\bm{0}$ (the zero vector in the parameter space).  This corresponds to the case where all the random vectors $V_1,\ldots,V_m,H_1,\ldots,H_n$ are iid with a symmetric Bernoulli$(1/2)$ distribution; see Theorem \ref{thm:two}.
\end{itemize}

@@

{\bf Proof of Theorem \ref{thm:one}.} We first establish the equivalence of claims 2 and 3. If claim 2 holds, then
 claim 3 holds trivially. Suppose claim 3 holds and pick $\ell \in \{1,\ldots,m\}$. Let  $\theta_{v_1},\ldots,\theta_{v_m},\theta_{h_1},\ldots,\theta_{h_n}$, $\theta_{ij}$, $i=1,\ldots,m$, $j=1,\ldots,n$, denote the components of $\bm{\theta}_1$ and let $\tilde{\theta}_{v_1},\ldots,\tilde{\theta}_{v_m},\tilde{\theta}_{h_1},\ldots,\tilde{\theta}_{h_n}$, $\tilde{\theta}_{ij}$, $i=1,\ldots,m$, $j=1,\ldots,n$, denote the analogous components of $\bm{\theta}_2$. Then, by assumption,
\begin{align*}
\exp\left[ 2\theta_{v_\ell} +2 \sum_{j=1}^n  h_j \theta_{\ell j} \right] &= \frac{P_1\left(\{V_\ell = 1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)} {P_1\left(\{V_\ell = -1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}\\
&= \frac{P_2\left(\{V_\ell = 1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}
{P_2\left(\{V_\ell = -1\} \cap \bigcap_{i=1 \atop i \neq \ell}^m \{V_i=v_i\} \cap \bigcap_{j=1}^n \{H_j=h_j\}\right)}\exp\left[ 2\tilde{\theta}_{v_\ell} +2 \sum_{j=1}^n  h_j \tilde{\theta}_{\ell j} \right]
\end{align*}
for any $\{v_i\}_{i=1,i\neq \ell}^m,\{h_j\}_{j=1}^n \subset\{\pm 1\}$. Equating exponents, we have $\theta_{v_\ell} - \tilde{\theta}_{v_\ell} = \sum_{j=1}^n  h_j (  \tilde{\theta}_{\ell j} -\theta_{\ell j})$ for any $\{h_j\}_{j=1}^n \subset \{\pm 1\}$. Taking all $h_j=1$ or all $h_j=-1$ for $1 \leq j \leq n$ yields $\theta_{v_\ell} = \tilde{\theta}_{v_\ell}$. This result further implies $\tilde{\theta}_{\ell 1} -\theta_{\ell 1} = -  \sum_{j=2}^n  h_j (  \tilde{\theta}_{\ell j} -\theta_{\ell j})$ for
any given $\{h_j\}_{j=2}^n \subset \{\pm 1\}$, so that $\theta_{\ell 1}=\tilde{\theta}_{\ell 1}$ must hold. Iterating this argument, sequentially for each $j=2,\ldots,m$, shows $\theta_{\ell j}=\tilde{\theta}_{\ell j}$ for any $1 \leq j \leq n$ in addition to $\theta_{v_\ell}=\tilde{\theta}_{v_\ell}$.  As $\ell\in\{1,\ldots,m\}$ was arbitrary, we have $\theta_{v_i}=\tilde{\theta}_{v_i}$ and $\theta_{i j}=\tilde{\theta}_{i j}$ for any $1 \leq i \leq m$, $1 \leq j \leq n$.

Now pick $k \in \{1,\ldots,n\}$ so that, by the RBM$(m,n,\bm{\theta})$ probability structure as above, we have analogously that
$$
\theta_{h_k} + \sum_{i=1}^m  v_i \theta_{i k} =  \tilde{\theta}_{h_k} + \sum_{i=1}^m  v_i \tilde{\theta}_{i k}
$$
for any $\{v_i\}_{i=1}^m \subset \{\pm 1\}$, so that $\theta_{h_k}  =  \tilde{\theta}_{h_k}$. As $k \in \{1,\ldots,n\}$ was arbitrary, we have now established claim 2 from claim 3.

To establish the equivalence of claims 1 and 3 in Theorem \ref{thm:one} under the RBM$(m,n,\bm{\theta})$ model, we note that claim 3 easily implies claim 1. Hence, it suffices now to establish claim 3 from claim 1. This follows immediately from Lemma \ref{lemma:one} below and the fact that the variables $V_1,\ldots,V_m$ are conditionally independent given $H_1,\ldots,H_n$ in the RBM$(m,n,\bm{\theta})$ model (and likewise $H_1,\ldots,H_n$ are independent given $V_1,\ldots,V_m$).

\begin{applemma} 
\label{lemma:one}
Let $m,n\geq 1$.  Let $P_1,P_2$ denote two probability distributions for a random vector $(\bm{V}_m,\bm{H}_n) = (V_1,\ldots,V_m, H_1,\ldots,H_n)$ supported on $\{a,b\}^{m+n}$, for some real constants $a\neq b \in\mathbb{R}$, such that $P_i( \bm{V}_m=\bm{v}_m, \bm{H}_n=\bm{h}_n)>0$, $i=1,2$, for any $\bm{v}_m \in \{a,b\}^{m}$, $\bm{h}_n \in \{a,b\}^{n}$. Suppose additionally that, under $P_1$ or $P_2$,  the variables $V_1,\ldots,V_m$ are conditionally independent given $\bm{H}_n=\bm{h}_n \in \{a,b\}^{n}$ and that the variables $H_1,\ldots,H_n$ are conditionally independent given $\bm{V}_m=\bm{v}_m \in \{a,b\}^{m}$.

### 2016-11-05 writing/draft.Rmd
reorganize dir and try to soften language about empirical likelihood

---
title: "On the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Daniel Nordman, Stephen Vardeman"
date: "September, 2016"
linestretch: 2 
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thereby thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs has been largely unexplored. In this paper, we discuss the relationship between RBM parameter specification in the binary case and the prevalence of undesirable model behaviors such as degeneracy, instability and/or uninterpretability. We also describe the difficulties that arise in likelihood-based and Bayes fitting  of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
---

@@

Additionally, because RBM models are highly overparametrized, any fully principled fitting method based on the likelihood will simply reproduce the (discrete) empirical distribution of most realistic training sets. That is, there is no "smoothed distribution" achieved in fitting a RBM model with a rigorous likelihood-based method. We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve prediction or inference in a principled way.

@@

However, computational concerns are not the only issues with fitting an RBM using ML. In addition, the RBM model, with the appropriate choice of parameters, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the $-1/1$ valued data outcomes). To see this, we provide a small example.

@@

The previous discussion intends to indicate that the RBM model class suffers from parameter identifiability issues that go beyond mere symmetries in the parametrization. Quite different parameter settings can induce the same essential RBM model. However, this  is not the most disastrous implication of the RBM parameterization. A far worse consequence is that, when fitting the RBM model by likelihood-based methods, we already know what the answer must be before we begin: namely, such  fitting will simply aim to reproduce the empirical distribution from the data. That is, based on a random sample of visible variables, the model for the cell probabilities that has the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the parametrization of the RBM model itself ensures that this empirical distribution can be arbitrarily well approximated.

@@

```{r child='appendix-center.Rmd'}
```

>>>>

---
title: "On the propriety of restricted Boltzmann machines"
author: "Andee Kaplan, Daniel Nordman, and Stephen Vardeman"
date: "November, 2016"
linestretch: 2 
output: 
  pdf_document:
    fig_caption: yes
    includes:
      in_header: latex/header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thereby thought to have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs has been largely unexplored. In this paper, we discuss the relationship between RBM parameter specification in the binary case and the prevalence of undesirable model behaviors such as degeneracy, instability and/or uninterpretability. We also describe the difficulties that arise in likelihood-based and Bayes fitting  of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
---

@@

For a RBM model with sufficient hidden variables, it has been shown that any distribution on the data can be approximated arbitrarily well [@le2008representational, @montufar2011refinements, and @montufar2011expressive]. The empirical distribution is the best fitting model to the observed cell data and so the RBM will seek to achieve this when possible. Thus, any fully principled fitting method based on the likelihood of a RBM with enough hidden variables will simply reproduce the (discrete) empirical distribution of most realistic training sets. That is, there is no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method. We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve prediction or inference in a principled way without intentially limiting the flexibility of the model.

@@

However, computational concerns are not the only issues with fitting an RBM using ML. In addition, the RBM model, with the appropriate choice of parameters and number of hiddens, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the $-1/1$ valued data outcomes). @montufar2011refinements show that any distribution on $\{0, 1\}^{\nv}$ can be approximated arbitrarily well by a RBM with $2^{\nv}/2 - 1$ hidden units. To see this, we provide a small example.

@@

The previous discussion intends to indicate that the RBM model class suffers from parameter identifiability issues that go beyond mere symmetries in the parametrization. Quite different parameter settings can induce the same essential RBM model. However, this is not the most disastrous implication of the RBM parameterization. A far worse consequence is that, when fitting the RBM model by likelihood-based methods, we already know what the answer must be before we begin: namely, such  fitting will simply aim to reproduce the empirical distribution from the data if sufficiently many hiddens are in the model. That is, based on a random sample of visible variables, the model for the cell probabilities that has the highest likelihood over  *all possible model classes* (i.e., RBM-based or not) is the empirical distribution, and the parametrization of the RBM model itself ensures that this empirical distribution can be arbitrarily well approximated.

@@

```{r child='appendix/appendix-center.Rmd'}
```
